Governing through patronage: the bargain for education in municipalities of Brazil

### 2022-08-31 paper/paper.tex
fix: fix compiling of paper

\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1.5in, bottom = 1.5in, left = 1.5in, right = 1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{hyperref}

@@

\title{Governing through patronage: the bargain for education in decentralized Brazil}
\date{\today}

\author{Galileu Kim}
\affil{Princeton University}

@@

\begin{abstract}
    How can governments improve the quality of education? I argue that local politicians govern through patronage, undermining education at the expense of voters. Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

\newpage
\section{Introduction}

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty". See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}. Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them. \citep{cochrane_policies_1983, falleti_decentralization_2010}. How can local governments improve the quality the of public services they deliver? In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can public education be improved?

Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions? Who holds power over public education, and whose interests is it meant to serve?

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

@@

In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}. In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction. The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

@@

Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
    \label{fig:staff_turnover}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
\end{figure}

I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$

\blue{Insert more information and descriptive paragraph on turnover index here}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

@@

Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.

\begin{proposition}
    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
\end{proposition}

Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity. Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v'}, W_B)$. If $k' = (n + 1)/2$, then we know that $k' \leq k^*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.

From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$. Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$. Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$. But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$. Therefore, the following holds:

\begin{proposition}
    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
\end{proposition}

In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases. Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive. Similarly

\begin{proposition}
    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
\end{proposition}

The ``convexity" of $E$ guarantees that local information is sufficient to generate comparative statistics regarding changes in preferences $\mathbf{v}' \rightarrow \mathbf{v}$. We can characterize the change in total expenditures as a result of a shift in voter preferences. Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$. From equation \ref{eqn:total_expenditure} we have

@@

Enacting policy requires the exchange of political currency for votes. Whether it be in presidential coalitionism, or in the local city council politics, mayors who wish to govern have to engage in transactions with the legislature. What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: more patronage occurs, leading to worse public service outcomes.

The model also highlights a key aspect of clientelism that is often neglected electoral accountability models: voters have a limited voice. Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and more to do with the city councilors. The first order requirement for the government is to ensure that it has enough legislative votes in order to enact the very policies that the voter may or may not desire. This transactional cost is not illegal: rather, it is necessary for democratic relations between different branches of government.

@@

\section{Empirical results}
\label{sec:empirics}

@@

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

For the first estimation, I test the hypothesis that mayors who control less seats in the city council engage in more patronage. To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates. I deploy three sets of models. For the second and third estimation I turn to the micro, bureaucrat-level and a then to aggregated turnover at the municipal level. The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year. The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year. For the second set of estimations, the main specification is:

@@

I first turn to testing the main proposition of this paper: that turnover in educational staff responds to the degree of political alignment between the mayor and politician. In line with theory, there is strong evidence that mayors who are electoral allies with more (less) seats in the city council resort to less (more) patronage. This is robust to a set of specifications, including state-year fixed effects and alternative measures of turnover in educational staff.

@@

Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE) in our estimations. Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores. We present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover.

@@

Finally, I present some evidence of differential returns to patronage, in particular for city councilors. For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term manda While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education. The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}

@@

\section{Conclusion}
\label{sec:conclusion}

@@

\section*{Appendix}

>>>>

% !TEX root = paper.tex
\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1.5in, bottom = 1.5in, left = 1.5in, right = 1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{hyperref}

@@

\title{Governing through patronage: the bargain for education in municipalities of Brazil}
\date{\today}

\author{Galileu Kim}
\affil{DIME, World Bank}

@@

\begin{abstract}
    How do political institutions shape decisions over bureaucratic personnel? And what are the implications of these management decisions on public service delivery? I argue that local mayors govern through patronage, lowering the quality of educational services. Qualitative evidence gathered in fieldwork in Brazil, as well as large-scale administrative data suggest that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering positions in the department of education. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

\section{Introduction}

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy remains widespread, basic mathematical operations inaccessible.\footnote{The World Bank calls this ``learning poverty". See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}. Politicians play an important role in deciding whether or not to deliver these services \citep{gulzar_politicians_2017,besley_political_2003}, but alongside elected officials, bureaucratic institutions shape how these services are delivered \citep{pepinsky_bureaucracy_2017, finan2017personnel}. How do politicians manage bureaucratic institutions, and what are their downstream consequences for public services?

Previous studies have found that individual-level incentives, electoral or performance related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on targeted interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But across the developing world, political and bureaucratic actors do not govern alone. Public officials are embedded in institutions that shape and constrain their actions \citep{thelen_historical_1999}. Moreover, it is common that instead of a single political actor, multiple principals compete over the same pool of resources, requiring coalitional bargains to effectively govern \citep{francois_how_2015}. How do these coalitional bargains affect patronage appointments, and what are its implications for the population they are meant to serve?

To answer these questions, I conducted fieldwork research in municipalities across Brazil over the course of two years. In interviews with local politicians, school principals and teachers, I uncovered the pervasive influence of clientelism, and how it extends to public education with negative consequences on educational outcomes \citep{stokes2013brokers}. Clientelism in the educational sector was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government that occur between elections \citep{raile_executive_2011}. In Brazilian municipalities, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

In this paper, I argue that public education in municipalities across Brazil is a bargaining chip between political elites, with personnel appointments into public schools and management dominated by transactions occurring between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over constituency welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The empirical identification of these theoretical mechanisms proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of model specifications and different measures of turnover.

@@

In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}. In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and subtraction. The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school students attend a public school. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

@@

Brazil collects fine-grained data on education and makes it publicly available for research. Data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both Mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown.pdf}
    \caption{Proportion of public sector jobs by department}
    \label{fig:staff_turnover}
\end{figure}

I propose three different specifications for measuring bureaucratic turnover. From the RAIS I calculate the municipal-level percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, grade $g$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

$$\text{turnover}_{sgjt} = \frac{\text{exit}_{sgjt} + \text{entry}_{sgjt}}{N_{sgjt} + N_{sgjt-1}}$$

The following example illustrates how the index operates. Suppose that in school $s$ for the year 2010, there are three teachers working. In the following year, one new teacher arrives and two leave. Our index would reflect these changes: $\text{turnover} = \frac{1 + 2}{3 + 2} = 0.6$. If, in contrast, no teachers left or entered the school, we would instead have a turnover index of 0 for the school. The added advantage of this measure is that it simultaneously captures a broader definition of bureaucratic turnover, that includes both exits and entries into the local school. Figure \ref{fig:hist_turnover} presents the distribution of turnover index over time.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/hist_turnover_index.pdf}
    \caption{Distribution of turnover index by year, pooled across schools. Note that for robustness, only schools with more than 5 teachers are included.}
    \label{fig:hist_turnover}
\end{figure}

Note that there is not significant variation over time of the turnover index, indicating that there are no visible time trends or reduction in teacher turnover. This indicates that turnover is a chronic challenge for schools in Brazil and has not reduced significantly over time, despite efforts by the Ministry of Education to increase professionalization of eduactional staff through more strict public exam requirements. It is important to highlight that entry or exit in this context does not indicate that a teacher is necessarily leaving or entering the educational system, but rather being relocated to another school within the same system.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the eletcoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

@@

% Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:deltAk} we have the following.

% \begin{proposition}
%     (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
% \end{proposition}

% Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity. Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v'}, W_B)$. If $k' = (n + 1)/2$, then we know that $k' \leq k^*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.

% From equation \ref{eqn:optimal_k} we infer that $\Delta(k' - 1, \mathbf{v}', W_B) < 0$, which from equations \ref{eqn:deltAk} and \ref{eqn:def_deltAk} is equivalent to $v_k' > T(k' - 1, W_B)$. Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$. Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$. Therefore, from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$. The following proposition thus holds:

% \begin{proposition}
%     For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
% \end{proposition}

% Substantively, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases. Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive.

With this theoretical framework, we can characterize change in total expenditures as a result of a shift in voter preferences. Given two preferences $\mathbf{v}$ and $\mathbf{v}'$, write $\mathbf{v}$ and $\mathbf{v}'$ if $v_i \geq v'_i$ for all $i \in N$. From equation \ref{eqn:total_expenditure} we have

@@

Enacting policy requires the exchange of political currency for votes. Whether it be in presidential coalitionism, or in the local city council politics, mayors who wish to govern have to engage in transactions with the legislature. What I showed in this section was that political misalignment between the government and the legislature can in fact be counterproductive: the more opposition in the legislature a mayor faces, the more patronage she has to engage in to coopt them.

The model also highlights a key aspect of clientelism that is often neglected electoral accountability models: voters have a limited voice. Ultimately, the exchanges which occur between the legislature and the mayor have little to do with the voter at the end of the pipeline, and far more to do with the city councilors. The first order requirement for the mayor is to ensure that she has enough legislative votes in order to enact the very policies that the voter may or may not desire. This transactional cost is not illegal by design. It is often the cost for democratic relations between different branches of government.

@@

\section{Empirical results}
\label{sec:empirics_edu}

@@

This study presents two sets of estimations. The first one estimates the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores. To avoid interference across units of analysis, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

For the second estimation, I test the hypothesis that mayors who control less seats in the city council engage in more patronage. To do so I leverage employment data from RAIS and school census data to measure teacher turnover rates. I deploy three sets of models. For the second and third estimation I turn to the micro, bureaucrat-level and then to aggregated turnover at the municipal level. The first model is a logistic regression where the outcome of interest is whether a particular teacher or school principal is hired/fired for any given year. The second model is a linear model with fixed effects, where the outcome of interest is the proportion of educational staff hired/fired for a muncipality at any given year. For the second set of estimations, the main specification is:

@@

I first turn to testing the main proposition of this paper: that turnover in educational staff responds to the degree of political alignment between the mayor and the legislature. In line with theory, there is strong evidence that mayors who are electoral allies with more (less) seats in the city council resort to less (more) patronage. This is robust to a set of specifications, including state-year fixed effects and alternative measures of turnover in educational staff.

@@

Moving on to characterizing the downstream effects of patronage, I find that teacher and school principal turnover have significant, negative effects on average test scores. These results are robust to alternative specifications of turnover, as well as the use of different quality metrics (SAEB, SPAECE). Table \ref{tbl:student_learning} presents the results of the estimation on SAEB and SPAECE test scores. I present two specifications for turnover. Turnover index measures the amount of turnover in teachers at the school level. Work experience for teachers and school principals serve as an alternative measure of turnover.

@@

Finally, I present some evidence of differential returns to patronage, in particular for city councilors. For each position (mayor, city councilor) I estimate the probability of being reelected conditional on the amount of patronage occurred in the first term mandate. While mayors themselves do not directly benefit from greater patronage, city councilors who are not members of the electoral base of the mayor seem to be negatively affected by patronage appointments into education. The negative consequences of patronage on student learning, however, do not bite, as nayors who deliver a worse quality of education are no less likely to be reelected in the next term.\footnote{See appendix \ref{app:accountability}.}

@@

\section{Conclusion}
\label{sec:conclusion_edu}

@@

\pagebreak
\section*{Appendix}
